Different forms of uncertainty differently affect the evolution of social learning

### 2022-08-13 ms.tex
final discussion revision
Updates from Overleaf

Social learning enhances problem solving when acquiring information from others is more
efficient than learning on one's own~\cite{Laland2004}. However, social learning can also
lead individuals astray if, for example, they are copying outdated information.
Theoretical models have helped clarify the circumstances under which social learning
should be evolutionarily favored ~\cite{BoydRicherson1985}. This theoretical work has
motivated much empirical research and helped predict social learning behavior in humans
and other species ~\cite{McElreath2005,Kendal2018,Allen2019}.  This literature
demonstrates that social learning is a critical adaptation found across taxa for dealing
with uncertainty.  Uncertainty weighs particularly heavily on human adaptation because of
our long lifespans, cosmopolitan distribution, and the resulting necessity to adapt to
relatively coarse-grain environments \cite{levins1962}. However, the term ``uncertainty"
often conflates several different interpretations of that word. For example, different
models define uncertainty in terms of the rate of environmental change, spatial
environmental variation, or the reliability of information acquired from the environment.
Furthermore, in translating mathematical models to verbal predictions, the formal meaning
of terms like uncertainty is often lost \cite{lawson1988probability}, facilitating such
conflation. Furthermore, it can be important to distinguish a narrower definition of
uncertainty wherein the probabilities of outcomes remain unknown~\cite{knight1921,
volz2012}. As the number of formal models of social learning have expanded, an ever
increasing number of modeling choices ~\cite[Figure 1]{Kendal2018} and formalizations of
uncertainty have made it difficult to compare across models and consolidate our
understanding of the contexts in which social learning is adaptive. 
% Finally, while risk and uncertainty are often used interchangeably to refer to
% non-deterministic outcomes,

@@

Despite the theoretical centrality of temporal environmental variability, this construct is
implemented in several ways~\cite{aoki2014evolution}. Most commonly, it is modeled
as an independent probability of the environment changing its state each
generation
~\cite{BoydRicherson1985,Rogers1988,Feldman1996,McElreath2005,Enquist2007,perreault2012bayesian,aoki2014evolution}.
% The probabilities of environmental change are usually fixed per generation
% ~\cite{BoydRicherson1985,Rogers1988,kendal2009evolution}, 
but has also been modeled as deterministic cycles~\cite{Feldman1996, aoki2014evolution}.
Furthermore, the consequences of environmental change can range from mild to
catastrophic. In the latter case, a change of environment results in the total
elimination of any adaptive behavior, which must be learned \emph{de novo} for the
new environment~\cite{Rogers1988}. Such catastrophic environmental changes present
a large adaptive challenge as individuals cannot rely on accumulated information
from previous generations. Other models of environmental change introduce less
uncertainty as they fluctuate between two environmental states with corresponding
adaptive behaviors. This means that previously maladaptive
behaviors become adaptive when the environment changes
~\cite{perreault2012bayesian}. The chosen mechanism for modelling temporal
change has real theoretical consequences~\cite{aoki2014evolution}.

\textbf{Selection set size.}
When one does not know which option to take, uncertainty increases with the number of options, which we call the selection set size.
% This is a mathematical fact at the
% core of information theory: the number of bits required to reliably encode the optimal solution increases by one every time the
% selection set doubles in size~\cite{mackay2003information}. 
% This is intuitive; the more options to consider, the greater one's uncertainty about which option is the best choice. 
In studies of social learning, the selection set is often limited
to just two options.  For example, in empirical studies of \emph{bombus
terresteris} (bumble bee)~\cite{Baracchi2018} and \emph{parus major} (great
tit)~\cite{Aplin2017}, experimenters provided two behaviors for the bees or birds
to choose from, where one yielded a higher payoff than the other. %respectively,
each yielding greater or smaller payoffs depending on experimental treatment.
Similarly, many human studies have used just two or three possible
behaviors~\cite{McElreath2005,Toyokawa2019}. Behavioral study designs are
often motivated by modeling studies with similar formulations~\cite{Rogers1988,boyd1995does,
perreault2012bayesian}.
% \cm{pretty sure boyd1995 uses same trick as
% rogers 88 so not sure why they were being used as illustrating separate kinds of
% behavior selection sets. correct if wrong} 
Larger, more open sets of behavioral
choices are not uncommon in experiments trying to capture more complex or
naturalistic tasks~\cite{derex2013, wasielewski2014}, but these map on
imperfectly to the theoretical literature that tend to use a limited set of behavioral options.
%\mt{How so?}. 
Some models have studied systems with
larger, but defined, selection sets~\cite{Rendell2010}, while in others the
number of options is implicit in the probability that a learner gets the right answer.
~\cite{Enquist2007} 
%\mt{How does this work exactly in Enquist et al 2007?}. 
Researchers have also compared models with a different number
of environment states, corresponding to a different number of optimal behaviors
through time ~\cite{Feldman1996}. These examples share one common feature: 
the size of the selection set was held static \emph{within} model generations.
% as a measure of changing uncertainty in cultural evolutionary models.

\textbf{Payoff ambiguity.} Many models of social learning differentiate between
the payoffs for adopting optimal vs. non-optimal behaviors
\cite{Rogers1988,Enquist2007,Rendell2010}. The size of the difference between
these payoffs is usually taken to indicate the strength of selection of learning,
which makes sense. However, in reality payoffs for particular behaviors are not
always consistent. A behavior may yield a large payoff sometimes and a small
payoff other times \cite{McElreath2005}. This means that signals about the
relationship between behavior and payoff are often noisy, and differentiating
between behavioral options is in part a problem of signal detection.  When the
difference in expected payoffs between optimal and non-optimal behaviors is very
large, this noise matters little, as the signal is still very clear. However, when
the expected payoffs of different behaviors are similar relative to the size of
their variances, ambiguity arises about which behaviors really are superior.
Smaller differences between payoffs corresponds to larger ambiguity. Payoff
ambiguity has been manipulated in both theoretical ~\cite{perreault2012bayesian}
and empirical ~\cite{McElreath2005, Morgan2012} studies, both of which support the
claim that payoff ambiguity increases the reliance on social information.
Importantly, payoff ambiguity %differs from selection set size in that it not only
affects uncertainty, it also directly influences both uncertainty \emph{and} the
strength of selection. %This parameter sometimes takes the form of differences in
expected payoffs arising from different behaviors or
strategies~\cite{Enquist2007,Rendell2010} or by varying the standard deviation of
payoffs~\cite{McElreath2005}.

@@

Empirically, the number of learning opportunities can be manipulated in the lab,
but will also correspond with an individual's age the real world. In multi-round
studies of information use in novel tasks, US participants' use of social
information declined precipitously across rounds ~\cite{McElreath2005}, suggesting
they were more likely to use social information when they were most uncertain
about the task early on. In a more naturalistic context, Aplin et al. (2017)
\nocite{Aplin2017} found that younger great tits more readily used social
information compared with older individuals, possibly because they had accrued
less information via individual learning, and also possibly because younger
individuals have the most to gain (given their expected remaining lifespan) by
switching to superior behavioral options. 
% \ps{I got rid of the Leris and Reader reference because it didn't seem relevant
% to age-based uncertainty. They just showed that young guppies exposed to
% reliable models were more likely to use social information later. I also added
% some human studies, which seems important.} 
Cross-cultural studies on humans have shown the importance of childhood as a phase
of heavy social learning ~\cite{Reyes2016}. Young children are more likely to
acquire their beliefs and simple skills from their parents than are older children
or adults, which is at least partly due to the differential knowledge accrual
between young children and the older adults to whom they direct the most trust and
attention \cite{kline2013teaching}.
%The age of individuals was found to have an effect on social learning in both
%\emph{parus major}~\cite{Aplin2017} and \emph{poecilia reticulata}
%(guppies)~\cite{Leris2016}. \ps{What effect? I'm not sure } \mt{Need better
%connection between lifespan and age, or better references to support use of
%lifespan}.
The effective lifespan for learning
varies across tasks, individuals, and species, yet most models assume one learning
opportunity per generation ~\cite{Feldman1996,Henrich1998, perreault2012bayesian}.
Some models do allow several cultural generations within 1 genetic generation
~\cite{Enquist2007}, but little formal theory explicitly examines the role of
learning opportunities in social learning evolution models.

@@

Social learning is the adaptive icing on the cake of other cognitive adaptations
to ecological uncertainty.  While the literature reviewed above suggests that
several forms of uncertainty play a role in favoring the evolution of social
learning, several other cognitive mechanisms have also evolved in response to
uncertainty ~\cite{volz2012}. Many of these are flexible learning mechanisms that
do not require assessing others' information. For example, people in experimental
studies increase their exploratory behavior when total uncertainty in the task
increases and are more likely to test %Furthermore, correlations in uncertainty
between potential behaviors, humans tend to adopt directed exploration strategies
that favor testing behaviors with greater observed payoff variance
\cite{Wilson2014,Gershman2019}.

@@

Our modeling choices were judiciously made to address the main question of how
uncertainty affects the evolution of social learning in general; however there exist
alternative empirically-valid choices that could help further develop a more nuanced
theory of social learning.  Our model necessarily made simplifying assumptions that
may not always hold in more specific situations, though they are empirically justified
to be general across taxa. First, note that we only considered success-biased
learning, although conformity is sometimes the dominant social learning
mechanism~\cite{Muthukrishna2016a,Smaldino2018b}.  Conformity may lead to more drift
overall, since success-biased teacher selection, which we assumed, maximizes the
potential benefit to social learners. We assumed that the number of behaviors and
their payoffs were constant, but this fails to account for evolutionary feedback which
creates new behavioral opportunities as time progresses, e.g.\ via niche
construction~\cite{Smaldino2012a,Heras-Escribano2020} or cumulative cultural evolution
supported by dynamic network structure~\cite{Smolla2019,Derex2020}.  Group structure
and processes such as homophily and in-group bias/out-group aversion could inhibit the
evolution of social learning since successful outgroup teachers could be cut off from
in-group learners~\cite{Jackson2012}. We assumed agents choose behaviors via softmax
search, but this is an underpowered algorithm compared to human individual
learning~\cite{Schulz2020a,Wu2022}.  More powerful individual learning could further
support social learning.  These considerations reveal new opportunities to further by
adjusting or extending the present model for other contexts.

>>>>

Social learning enhances problem solving when acquiring information from others is more
efficient than learning on one's own~\cite{Laland2004}. However, social learning can also
lead individuals astray if, for example, they are copying outdated information.
Theoretical models have helped clarify the circumstances under which social learning
should be evolutionarily favored~\cite{BoydRicherson1985}. This theoretical work has
motivated much empirical research and helped predict social learning behavior in humans
and other species~\cite{McElreath2005,Kendal2018,Allen2019}.  This literature
demonstrates that social learning is a critical adaptation found across taxa for dealing
with uncertainty.  Uncertainty weighs particularly heavily on human adaptation because of
our long lifespans, cosmopolitan distribution, and the resulting necessity to adapt to
relatively coarse-grain environments \cite{levins1962}. However, the term ``uncertainty"
often conflates several different interpretations of that word. For example, different
models define uncertainty in terms of the rate of environmental change, spatial
environmental variation, or the reliability of information acquired from the environment.
Furthermore, in translating mathematical models to verbal predictions, the formal meaning
of terms like uncertainty is often lost \cite{lawson1988probability}, facilitating such
conflation. Furthermore, it can be important to distinguish a narrower definition of
uncertainty wherein the probabilities of outcomes remain unknown~\cite{knight1921,
volz2012}. As the number of formal models of social learning have expanded, an ever
increasing number of modeling choices~\cite[Figure 1]{Kendal2018} and formalizations of
uncertainty have made it difficult to compare across models and consolidate our
understanding of the contexts in which social learning is adaptive. 
% Finally, while risk and uncertainty are often used interchangeably to refer to
% non-deterministic outcomes,

@@

Despite the theoretical centrality of temporal environmental variability, this construct is
implemented in several ways~\cite{aoki2014evolution}. Most commonly, it is modeled
as an independent probability of the environment changing its state each
generation
~\cite{BoydRicherson1985,Rogers1988,Feldman1996,McElreath2005,Enquist2007,perreault2012bayesian,aoki2014evolution}.
% The probabilities of environmental change are usually fixed per generation
%~\cite{BoydRicherson1985,Rogers1988,kendal2009evolution}, 
but has also been modeled as deterministic cycles~\cite{Feldman1996, aoki2014evolution}.
Furthermore, the consequences of environmental change can range from mild to
catastrophic. In the latter case, a change of environment results in the total
elimination of any adaptive behavior, which must be learned \emph{de novo} for the
new environment~\cite{Rogers1988}. Such catastrophic environmental changes present
a large adaptive challenge as individuals cannot rely on accumulated information
from previous generations. Other models of environmental change introduce less
uncertainty as they fluctuate between two environmental states with corresponding
adaptive behaviors. This means that previously maladaptive
behaviors become adaptive when the environment changes
~\cite{perreault2012bayesian}. The chosen mechanism for modelling temporal
change has real theoretical consequences~\cite{aoki2014evolution}.

\textbf{Selection set size.}
When one does not know which option to take, uncertainty increases with the number of options, which we call the selection set size.
% This is a mathematical fact at the
% core of information theory: the number of bits required to reliably encode the optimal solution increases by one every time the
% selection set doubles in size~\cite{mackay2003information}. 
% This is intuitive; the more options to consider, the greater one's uncertainty about which option is the best choice. 
In studies of social learning, the selection set is often limited
to just two options.  For example, in empirical studies of \emph{bombus
terresteris} (bumble bee)~\cite{Baracchi2018} and \emph{parus major} (great
tit)~\cite{Aplin2017}, experimenters provided two behaviors for the bees or birds
to choose from, where one yielded a higher payoff than the other. %respectively,
each yielding greater or smaller payoffs depending on experimental treatment.
Similarly, many human studies have used just two or three possible
behaviors~\cite{McElreath2005,Toyokawa2019}. Behavioral study designs are
often motivated by modeling studies with similar formulations~\cite{Rogers1988,boyd1995does,
perreault2012bayesian}.
% \cm{pretty sure boyd1995 uses same trick as
% rogers 88 so not sure why they were being used as illustrating separate kinds of
% behavior selection sets. correct if wrong} 
Larger, more open sets of behavioral
choices are not uncommon in experiments trying to capture more complex or
naturalistic tasks~\cite{derex2013, wasielewski2014}, but these map on
imperfectly to the theoretical literature that tend to use a limited set of behavioral options.
%\mt{How so?}. 
Some models have studied systems with
larger, but defined, selection sets~\cite{Rendell2010}, while in others the
number of options is implicit in the probability that a learner gets the right answer~\cite{Enquist2007}.
%\mt{How does this work exactly in Enquist et al 2007?}. 
Researchers have also compared models with a different number
of environment states, corresponding to a different number of optimal behaviors
through time~\cite{Feldman1996}. These examples share one common feature: 
the size of the selection set was held static \emph{within} model generations.
% as a measure of changing uncertainty in cultural evolutionary models.

\textbf{Payoff ambiguity.} Many models of social learning differentiate between
the payoffs for adopting optimal vs. non-optimal behaviors
\cite{Rogers1988,Enquist2007,Rendell2010}. The size of the difference between
these payoffs is usually taken to indicate the strength of selection of learning,
which makes sense. However, in reality payoffs for particular behaviors are not
always consistent. A behavior may yield a large payoff sometimes and a small
payoff other times \cite{McElreath2005}. This means that signals about the
relationship between behavior and payoff are often noisy, and differentiating
between behavioral options is in part a problem of signal detection.  When the
difference in expected payoffs between optimal and non-optimal behaviors is very
large, this noise matters little, as the signal is still very clear. However, when
the expected payoffs of different behaviors are similar relative to the size of
their variances, ambiguity arises about which behaviors really are superior.
Smaller differences between payoffs corresponds to larger ambiguity. Payoff
ambiguity has been manipulated in both theoretical~\cite{perreault2012bayesian}
and empirical~\cite{McElreath2005, Morgan2012} studies, both of which support the
claim that payoff ambiguity increases the reliance on social information.
Importantly, payoff ambiguity %differs from selection set size in that it not only
affects uncertainty, it also directly influences both uncertainty \emph{and} the
strength of selection. %This parameter sometimes takes the form of differences in
expected payoffs arising from different behaviors or
strategies~\cite{Enquist2007,Rendell2010} or by varying the standard deviation of
payoffs~\cite{McElreath2005}.

@@

Empirically, the number of learning opportunities can be manipulated in the lab,
but will also correspond with an individual's age the real world. In multi-round
studies of information use in novel tasks, US participants' use of social
information declined precipitously across rounds~\cite{McElreath2005}, suggesting
they were more likely to use social information when they were most uncertain
about the task early on. In a more naturalistic context, Aplin et al. (2017)
\nocite{Aplin2017} found that younger great tits more readily used social
information compared with older individuals, possibly because they had accrued
less information via individual learning, and also possibly because younger
individuals have the most to gain (given their expected remaining lifespan) by
switching to superior behavioral options. 
% \ps{I got rid of the Leris and Reader reference because it didn't seem relevant
% to age-based uncertainty. They just showed that young guppies exposed to
% reliable models were more likely to use social information later. I also added
% some human studies, which seems important.} 
Cross-cultural studies on humans have shown the importance of childhood as a phase
of heavy social learning~\cite{Reyes2016}. Young children are more likely to
acquire their beliefs and simple skills from their parents than are older children
or adults, which is at least partly due to the differential knowledge accrual
between young children and the older adults to whom they direct the most trust and
attention \cite{kline2013teaching}.
%The age of individuals was found to have an effect on social learning in both
%\emph{parus major}~\cite{Aplin2017} and \emph{poecilia reticulata}
%(guppies)~\cite{Leris2016}. \ps{What effect? I'm not sure } \mt{Need better
%connection between lifespan and age, or better references to support use of
%lifespan}.
The effective lifespan for learning
varies across tasks, individuals, and species, yet most models assume one learning
opportunity per generation~\cite{Feldman1996,Henrich1998, perreault2012bayesian}.
Some models do allow several cultural generations within 1 genetic generation
~\cite{Enquist2007}, but little formal theory explicitly examines the role of
learning opportunities in social learning evolution models.

@@

Social learning is the adaptive icing on the cake of other cognitive adaptations
to ecological uncertainty.  While the literature reviewed above suggests that
several forms of uncertainty play a role in favoring the evolution of social
learning, several other cognitive mechanisms have also evolved in response to
uncertainty~\cite{volz2012}. Many of these are flexible learning mechanisms that
do not require assessing others' information. For example, people in experimental
studies increase their exploratory behavior when total uncertainty in the task
increases and are more likely to test %Furthermore, correlations in uncertainty
between potential behaviors, humans tend to adopt directed exploration strategies
that favor testing behaviors with greater observed payoff variance
\cite{Wilson2014,Gershman2019}.

@@

Our modeling choices were judiciously made to address the main question of how
uncertainty affects the evolution of social learning in general; however there exist
alternative empirically-valid choices that could help further develop a more nuanced
theory of social learning.  Our model necessarily made simplifying assumptions that
may not always hold in more specific situations, though they are empirically justified
to be general across taxa. First, note that we only considered success-biased
learning, although conformity is sometimes the dominant social learning
mechanism~\cite{Muthukrishna2016a,Smaldino2018b}.  Conformity may lead to more drift
overall, since success-biased teacher selection, which we assumed, maximizes the
potential benefit to social learners. We assumed that the number of behaviors and
their payoffs were constant, but this fails to account for evolutionary feedback which
creates new behavioral opportunities as time progresses, e.g.\ via niche
construction~\cite{Smaldino2012a,Heras-Escribano2020} or cumulative cultural evolution
supported by dynamic network structure~\cite{Smolla2019,Derex2020}.  Group structure
and processes such as homophily and in-group bias/out-group aversion could inhibit the
evolution of social learning since successful outgroup teachers could be cut off from
in-group learners~\cite{Jackson2012}. We assumed agents choose behaviors via softmax
search, but this is an underpowered algorithm compared to human individual
learning~\cite{Schulz2020a,Wu2022}.  More powerful individual learning could further
support social learning.  These considerations reveal new opportunities to enhance
theoretical nuance by adjusting or extending the present model for other contexts.
